---
title: "AQS 88101 Tutorial 1: Explore AQS Data"
author: "Jonathan Callahan"
date: "September 29, 2021"
output:
  html_document:
    toc: TRUE
    toc_depth: 3
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This tutorial demonstrates how to use the **MazamaLocationUtils** and
**AirMonitorIngest** R packages to work with EPA AQS PM2.5 data. Target
audiences include grad students, researchers and any member of the public
interested in air quality and comfortable working with R and RStudio.

Tutorials in this series include:

* AQS 88101 Tutorial 1: Explore AQS Data
* AQS 88101 Tutorial 2: Create 'known locations' Table

# Background

The USFS [AirFire](https://www.airfire.org) group is focused on air quality 
measurements associated with wildfire smoke and maintains both historical and 
real-time databases of PM2.5 monitoring data obtained from stationary monitors. 
This data is used in operational displays and for retrospective analysis. Data 
ingest and management of air quality "stationary time series" are both important 
ongoing activities.

Challenges include:

* Location stability -- jitter in reported longitude and latitude coming from real-time GPS.
* Data duplication -- data from an individual monitor arriving through mutiple data streams.
* CPU-intensive spatial searches for data such as time zone.
 
To address these and other issues, we are creating "known locations" tables -- 
a set of locations which have been vetted for reuse and for which all 
CPU-intensive spatial queries have already been performed.

Separating purely spatial information from other monitor-related metadata will 
allow us to quickly assign incoming data to a "known location" and immediately 
retrieve associated spatial data that would otherwise require expensive spatial 
calculations.

The basic concept for creating a _"known locations"_ table is described in the 
introduction to the 
[MazamaLocationUtils](https://mazamascience.github.io/MazamaLocationUtils/)
R package.

For our initial "known locations" table, we will focus on AQS parameter 88101
(PM2.5 FRM/FEM) and measurement scales at the "NEIGHBORHOOD" scale and higher 
(>= 100 m), ignoring any "MICROSCALE" measurements found in the EPA AQS database. 
An initial set of "known locations" will be created from EPA AQS "sites" and 
"monitors" metadata tables and then standardized and augmented with additional 
spatial data.

# Setup

```{r mazama_setup, message = FALSE}
library(MazamaCoreUtils)

library(AirMonitorIngest)

# Create a directory specifically for EPA data
dir.create("~/Data/EPA", showWarnings = FALSE, recursive = TRUE)
```

We will begin with existing tables of EPA AQS site and monitor information. These 
are assumed to be reasonably well vetted and contain most of the relevant 
spatial information we will need for further data processing and visualization.

# Download Data

## AQS Sites

The `AQS_sites` table contains primarily spatial location information with additional
_"site"_ information that is specific to an individual monitor, _e.g._
`Owning Agency`. As we will see, it is possible to have multiple _"sites"_ at a 
single _"known location"_.

```{r AQS_sites, message = FALSE, results = "hold"}
# Get site metadata
AQS_sites <- epa_aqs_getSites(downloadDir = "~/Data/EPA")
```

The `AQS_sites` table contains `r nrow(AQS_sites)` records, each with `r ncol(AQS_sites)`
pieces of metadata:

```{r names_AQS_sites, display = FALSE}
names(AQS_sites)
```

## AQS Monitors

The `AQS_monitors` table mixes monitor-specific information like `Parameter Code`
and `Last Sample Date` with location-specific information like `State Code` or
`CBSA Name`.

```{r AQS_monitors, message = FALSE, results = "hold"}
# Get monitor metadata
AQS_monitors <- epa_aqs_getMonitors(downloadDir = "~/Data/EPA")
```

The `AQS_sites` table contains `r nrow(AQS_monitors)` records, each with `r ncol(AQS_monitors)`
pieces of metadata:

```{r names_AQS_monitors, display = FALSE}
names(AQS_monitors)
```

## AQS Parameter Codes

The `AQS_parameterCodes` table provides detailed information on each individual
parameter (pollutant) found in the AQS data.

```{r AQS_parameterCodes, message = FALSE, results = "hold"}
# Get parameter codes
AQS_parameterCodes <- epa_aqs_getCodes(tableName = "parameters")
```


The `AQS_parameterCodes` table contains `r nrow(AQS_parameterCodes)` records, each with `r ncol(AQS_parameterCodes)`
pieces of metadata:

```{r names_AQS_parameterCodes, display = FALSE}
names(AQS_parameterCodes)
```

Parameter code 88101 is associated with regulatory PM2.5 measurements.

_**NOTE:** Measured PM2.5 values are truncated. This will be relevant when storing data._

```{r parameter_88101, results = "hold"}
AQS_parameterCodes %>%
  dplyr::filter(`Parameter Code` == "88101") %>% 
  dplyr::glimpse()
```

----

# Subset for Sites with Parameter Code 88101

## Subset Monitors

We begin by subsetting `AQS_monitors` to include only those monitors used in
regulatory decision making -- PM2.5 FRM/FEM or `88101`. We will use state,
county and site information to create an `AQSID` which is typically associated
with a monitor-site. The `AQSID can be created for both `AQS_monitors` and
`AQS_sites`, allowing us to create queries combining the two.

```{r AQS_monitors_88101, results = "hold"}
AQS_monitors_88101 <-
  AQS_monitors %>%
  # Subset
  dplyr::filter(`Parameter Code` == "88101") %>%
  # Add AQSID as a useful unique identifier
  dplyr::mutate(
    AQSID = paste0(`State Code`, `County Code`, `Site Number`)
  )

dim(AQS_monitors_88101)
```

## Subset Sites

```{r AQS_sites_88101, results = "hold"}
AQS_sites_88101 <-
  AQS_sites %>%
  # Add AQSID as a useful unique identifier
  dplyr::mutate(
    AQSID = paste0(`State Code`, `County Code`, `Site Number`)
  ) %>%
  # Subset
  dplyr::filter(AQSID %in% AQS_monitors_88101$AQSID)

dim(AQS_sites_88101)
```

There are `r nrow(AQS_monitors_88101)` monitors but only `r nrow(AQS_sites_88101)`
sites. This is as it should be as monitors with different `POC` values may 
appear at the same site:

> This is the “Parameter Occurrence Code” used to distinguish different 
> instruments that measure the same parameter at the same site.

# Review Sites and Monitors

## Visual Inspection

We can create an interactive map showing the locations of the monitors and 
sites we have identified with monitors shown as red circles and sites shown as 
black rings. As shown below, we can visually review that every monitor is
associated with a site and _vise versa_ -- all circles have black rings around 
them and all black rings are filled with red.

Locations that appear darker red are evidence of multiple monitors (associated
with different `POC` values) at a single site.

```{r sites_monitors_leaflet}
monitor_locations <-
  AQS_monitors_88101 %>%
  dplyr::mutate(
    longitude = Longitude,
    latitude = Latitude
  ) %>%
  dplyr::mutate(
    locationID = MazamaLocationUtils::location_createID(longitude, latitude)
  )

site_locations <-
  AQS_sites_88101 %>%
  dplyr::mutate(
    longitude = Longitude,
    latitude = Latitude
  ) %>%
  dplyr::mutate(
    locationID = MazamaLocationUtils::location_createID(longitude, latitude)
  )

monitor_locations %>%
  MazamaLocationUtils::table_leaflet(
    locationOnly = TRUE,
    opacity = 0.0,
    fillColor = "red"
  ) %>%
  MazamaLocationUtils::table_leafletAdd(
    site_locations, 
    locationOnly = TRUE, 
    weight = 2,
    color = "black",
    fillOpacity = 0.0
  )
```

----

This ends the exploration of available data. 

See step 2 for the creation of a "known sites" table for these locations.

