---
title: "AQS 88101 Tutorial 3: Create Monitor 'meta' Table"
author: "Jonathan Callahan"
date: "September 29, 2021"
output:
  html_document:
    toc: TRUE
    toc_depth: 3
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This tutorial demonstrates how to use the **MazamaLocationUtils** and
**AirMonitorIngest** R packages to work with EPA AQS PM2.5 data. Target
audiences include grad students, researchers and any member of the public
interested in air quality and comfortable working with R and RStudio.

Tutorials in this series include:

* AQS 88101 Tutorial 1: Explore AQS Data
* AQS 88101 Tutorial 2: Create 'known locations' Table
* AQS 88101 Tutorial 3: Create Monitor 'meta' Table

# Setup

In the second tutorial we created a vetted "known locations" table for sites 
associated with parameter 88101.

We need to pick up where we left off in step 2:

```{r setup_data, message = FALSE}
library(MazamaCoreUtils)
library(AirMonitorIngest)

AQS_monitors <- epa_aqs_getMonitors(downloadDir = "~/Data/EPA")

AQS_monitors_88101 <-
  AQS_monitors %>%
  # Subset
  dplyr::filter(`Parameter Code` == "88101") %>%
  # Add AQSID as a unique identifier
  dplyr::mutate(
    AQSID = paste0(`State Code`, `County Code`, `Site Number`)
  )

library(MazamaSpatialUtils)
library(MazamaLocationUtils)

# Set up spatial data
MazamaLocationUtils::mazama_initialize("~/Data/Spatial")

# Set "known locations" directory
MazamaLocationUtils::setLocationDataDir("~/Data/known_locations")

sites_locationTbl <- MazamaLocationUtils::table_load("AQS_88101_sites")
```

# Validate Tables

Before we begin creating a 'meta' table for 88101 monitors, we need to validate
that the information we begin with is consistent. The following interactive
session shows that things are in order.

```{r validate_tablesl}
# Any AQSIDs in sites that are unknown in monitors?
setdiff(sites_locationTbl$AQS_AQSID, AQS_monitors_88101$AQSID)

# Any AQSIDs in monitors that are unknown in sites?
setdiff(AQS_monitors_88101$AQSID, sites_locationTbl$AQS_AQSID)

# NOTE:  This is the Wash. U. site we dealt with in Tutorial 2. This is why we
# NOTE:  cannot use 'AQSID' as an identifier for unique locations.

# Now create and compare locationIDs

AQS_monitors_88101$locationID <- 
  MazamaLocationUtils::location_createID(
    AQS_monitors_88101$Longitude, 
    AQS_monitors_88101$Latitude
  )

# Any locationIDs in sites that are unknown in monitors?
setdiff(sites_locationTbl$locationID, AQS_monitors_88101$locationID)

# Any AQSIDs in monitors that are unknown in sites?
setdiff(AQS_monitors_88101$locationID, sites_locationTbl$locationID)

# NOTE:  No differences means that 'locationID' is a unique identifier for
# NOTE:  unique locations.

# How many duplicate locations are found in monitors?
indices <- which(duplicated(AQS_monitors_88101$locationID))
length(indices)

# What 'POC's are they associated with.
AQS_monitors_88101 %>% 
  dplyr::slice(indices) %>% 
  dplyr::pull(POC) %>% 
  unique() %>% as.numeric() %>% sort()

# What 'POC's are found among monitors that have no co-habitants at their location
inverse_indices <- which(!duplicated(AQS_monitors_88101$locationID))
AQS_monitors_88101 %>% 
  dplyr::slice(inverse_indices) %>% 
  dplyr::pull(POC) %>% 
  unique() %>% as.numeric() %>% sort()
```

Everything looks in order. The `duplicated()` function only finds _subsequent_
instances and the `AQS_monitors_88101` appears to be organized so that `POC` of
"1" always comes before other `POC` values at a particular site.

As a reminder, 'POC' is described as:

> This is the “Parameter Occurrence Code” used to distinguish different 
> instruments that measure the same parameter at the same site.

# Merge Sites and Monitors

In creating a `meta` table for Monitor metadata, we wish to merge location-specific
information from `sites_locationTbl` with additional monitor-specific information
from `AQS_monitors_88101`.

Required columns include:

* deviceDeploymentID -- unique identifier (see MazmaLocationUtils)
* deviceID -- device identifier
* locationID -- location identifier (see MazmaLocationUtils)
* siteName -- English language name
* longitude -- decimal degrees E
* latitude -- decimal degrees N
* elevation -- elevation of station in m
* countryCode -- ISO 3166-1 alpha-2
* stateCode -- ISO 3166-2 alpha-2
* timezone -- Olson time zone

Additional columns may include anything relevant to the monitors or questions
being asked of the data.

## Create Tables to Join

Our goal is to join information from `sites_locationTbl` onto information from
`AQS_monitors_88101`. 

We begin by preparing `monitors_meta` to exclude any _site_ information found in
`AQS_monitors_88101` and to include the `locationID`:

```{r prepare_monitors_meta}
# Create core monitors_meta table

monitors_meta <-
  AQS_monitors_88101 %>%
  
  # Rename all existing columns with "AQS_"
  dplyr::rename_all(make.names) %>%
  dplyr::rename_all(~ gsub("^", "AQS_", .x))

# Keep the following columns
keep_columns <- c(
  # "AQS_State.Code",
  # "AQS_County.Code",
  # "AQS_Site.Number",
  "AQS_Parameter.Code",
  "AQS_Parameter.Name",
  "AQS_POC",
  # "AQS_Latitude",
  # "AQS_Longitude",
  # "AQS_Datum",
  "AQS_First.Year.of.Data",
  "AQS_Last.Sample.Date",
  "AQS_Monitor.Type",
  "AQS_Networks",
  "AQS_Reporting.Agency",
  "AQS_PQAO",
  "AQS_Collecting.Agency",
  "AQS_Exclusions",
  "AQS_Monitoring.Objective",
  "AQS_Last.Method.Code",
  "AQS_Last.Method",
  "AQS_Measurement.Scale",
  "AQS_Measurement.Scale.Definition",
  "AQS_NAAQS.Primary.Monitor",
  "AQS_QA.Primary.Monitor",
  # "AQS_Local.Site.Name",
  # "AQS_Address",
  # "AQS_State.Name",
  # "AQS_County.Name",
  # "AQS_City.Name",
  # "AQS_CBSA.Name",
  # "AQS_Tribe.Name",
  # "AQS_Extraction.Date",
  "AQS_AQSID",
  "AQS_locationID"
)

monitors_meta <- 
  monitors_meta %>%
  dplyr::select(dplyr::all_of(keep_columns))

# Check which columns are still found in both
intersect(names(monitors_meta), names(sites_locationTbl))

# Add additional variables

monitors_meta <-
  monitors_meta %>%
  # Include 'locationID' so we have something to join on
  dplyr::rename(
    locationID = AQS_locationID,
  ) %>%
  # Unique instrument ID
  dplyr::mutate(
    deviceID = sprintf("%s_%02d", AQS_AQSID, as.numeric(AQS_POC))
  ) %>%
  # Unique "device deployment" ID
  dplyr::mutate(
    deviceDeploymentID = paste(locationID, deviceID, sep = "_")
  ) %>%
  # Now remove AQS_AQSID
  dplyr::select(- AQS_AQSID)

# Check which columns are still found in both
intersect(names(monitors_meta), names(sites_locationTbl))
```

Now we can left join "known location" information to create a `meta` tibble:

```{r left_join_sites}
meta <-
  dplyr::left_join(monitors_meta, sites_locationTbl, by = "locationID")

# Reorder columns
coreNames <- c(
  "deviceDeploymentID", "deviceID",
  MazamaLocationUtils::coreMetadataNames
)
aqsNames <- setdiff(names(meta), coreNames)

meta <- 
  meta %>%
  dplyr::select(dplyr::all_of(c(coreNames, aqsNames)))
```

## Save our work

We will save our newly vetted "meta" table in a dedicated directory.

```{r save}
save(meta, file = ("~/Data/EPA/AQS_88101_meta.rda"))

readr::write_csv(meta, file = file.path("~/Data/EPA", "AQS_88101_monitors.csv"))
```


