---
title: "Working with WRCC Data"
author: "Jonathan Callahan"
date: "3/1/2022"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This article describes how to download and work with raw data from WRCC.
Begin by loading the **MazamaCoreUtils** and **AirMonitorIngest** libraries:

```{r libraries, echo = TRUE, eval=TRUE, warning=FALSE, message=FALSE}
library(MazamaCoreUtils)
library(AirMonitorIngest)

# Generate logging statements using MazamaCoreUtils logging functionality
logger.setLevel(TRACE)
```

## Raw data

The `wrcc_downloadData()` function sends a request to WRCC, asking for raw, 
ASCII data for a single monitor over some time period. 

```{r wrcc_downloadData}
fileString <- wrcc_downloadData(20210701, 20210930, unitID = 'sm16')
```

This `fileString` is just the tab-delimited data file and will look something
like this:

<pre>
 Smoke #16 
:       GMT	 Deg 	 Deg 	     	ser #	ug/m3	 Unk 	 l/m 	Deg C	  %  	 Unk 	deg C	  %  	 m/s 	 Deg 	volts	     
: Date/Time	  GPS  	  GPS  	Type   	Serial 	Conc   	 Misc  	 Ave.  	 Av Air	  Rel  	 Misc  	Sensor 	Sensor 	  Wind 	 Wind  	Battery	Alarm  
:YYMMDDhhmm	  Lat. 	  Lon. 	       	Number 	Hly Av 	  #1   	Air Flw	  Temp 	Humidty	  #2   	Int AT 	Int RH 	  Speed	 Direc 	Voltage	       
2107242200	39.714348	-105.123962	9	871600	11	-9999	2	31.4	32	83177	-9999	29	1	180	14	0
2107242300	39.714348	-105.123962	9	871600	18	-9999	2	32.5	30	83158	-9999	26	0.9	180	14	0
2107250000	39.71447	-105.124001	9	871600	26	-9999	2	31	32	83138	-9999	26	1.4	180	14	0
...
</pre>

The `wrcc_parseData()` function identifies the specific data format received and
properly parses the `fileString` into a tibble that is ready for examination.

```{r wrcc_parseData, results = "hold"}
tbl <- wrcc_parseData(fileString)

dplyr::glimpse(tbl)
```

At this point, you can create a time axis and plot the PM2.5 values. (Note that
the DateTime uses 2-digit years so we need to prepend `"20"` before parsing.)

```{r time_series}
time <- 
  paste0("20", tbl$DateTime) %>%
  MazamaCoreUtils::parseDatetime(timezone = "UTC")

plot(time, tbl$ConcRT, ylab = "PM2.5")
```

## Quality Control

QC algorithms are used to determine whether the reported PM2.5 values can be 
trusted.  The `wrcc_EBAMQualityControl()` and `wrcc_ESAMQualityControl()` functions
remove records whenever any parameters, _e.g._ flow rate, are out of their valid range.

When passed a tibble of raw data from WRCC, the `wrcc_qualityControl()` function 
will choose the proper QC algorithm and validate values using manufacturer
recommended thresholds. If logging is enabled, output will be printed describing
the QC process.

```{r QC}
tbl_qc <- wrcc_qualityControl(tbl)
```

Note that the number of rows is now reduced compared to the raw data.

```{r dim, results = "hold"}
dim(tbl)
dim(tbl_qc)
```

We can see which records were removed by plotting the raw data in red and
then overplotting the QC'ed data values in black.

```{r invalidated_values, results = "hold"}
plot(time, tbl$ConcRT, pch = 16, col = "red", ylab = "PM2.5")

# Need a new time axis for tbl_qc
time_qc <- 
  paste0("20", tbl_qc$DateTime) %>%
  MazamaCoreUtils::parseDatetime(timezone = "UTC")

points(time_qc, tbl_qc$ConcRT, pch = 16, col = "black")
```

----

Hopefully, this is enough to get you started working with WRCC raw data.
